---
title: "Data Simulation with Monte Carlo Methods"
author: "Marko Bachl"
institute: "University of Hohenheim"
format:
  revealjs:
      embed-resources: true
      height: 900
      width: 1600
      theme: [theme.scss]
      highlight-style: github-dark
      slide-number: c
      code-line-numbers: false
      history: false
      smaller: false
      title-slide-attributes:
        data-background-image: https://upload.wikimedia.org/wikipedia/commons/6/64/Uni_Hohenheim-Logo.svg
        data-background-position: "top right"
        data-background-size: auto
css: colors.css
---

```{r setup, include=FALSE}
pacman::p_load(knitr, extraDistr, tictoc, tidyverse)
set.seed(376)
```


# Hello there :)

## Introduction (I)

::: {.incremental}

- Hi, I'm Marko.

&nbsp;

-   I am not a statistician, econometrician, psychometrician, or any kind of \*ician.

-   Most importantly, I am also not a mathematician --- this is why I often need simulation methods.

-   I am also not a computer scientist or trained programmer.

:::

## Introduction (II)

-   I am a trained social scientist / communication researcher with an interest in quantitative methods.

-   I use data simulation methods for teaching myself and others as well as in my applied and methods research.

## Introduction (III)

### Random sample in *R*

```{r, echo=TRUE}
sample(x = 30, size = 4)
```

-   Who are you --- and why are you here?


## Traktandenliste

1)  Introduction & overview

2)  Monte Carlo Simulation?

3)  Proof by simulation --- the Central Limit Theorem (CLT)

4)  Errors and power --- torturing the *t*-test

5)  Misclassification and bias --- Messages mismeasured

6)  Outlook: What's next?


## Overview (I)

-   Introduction to Monte Carlo Simulation Methods
    -   How to think about simulation experiments
    -   How to get started in *R*

&nbsp;

-   How we will cover applied examples
    -   Simple, readable code, mostly `{tidyverse}`
    -   Data simulation "from scratch"


## Overview (II)

-   What we will *not* cover
    -   Packages for simulation experiments and data simulation (easier to use, but harder to understand)
    -   Advanced computing and programming stuff (not that hard to understand until its very hard to understand, very different skill set)
-   Some resources in the end

## Workshop concept

- Me talking: Lecture with code illustrations

- You talking: Questions

- All talking: Group exercices in breakout rooms

## Resources

Slides, scripts, and exercises:

-   [https://github.com/bachl/sim](https://github.com/bachl/sim)

R & packages:
```{r}
#| label: packages
#| output: "asis"
tibble(Package = c("R", sort(c(pacman::p_loaded())))) %>% 
  mutate(Version = map(Package, ~ as.character(pacman::p_ver(.x)))) %>% 
  str_glue_data("{Package} (Vers. {Version})") %>% 
  str_c(collapse = ", ") %>% 
  cat()
```


# Questions?
