---
title: "4a) Errors and power --- Torturing the *t*-test"
subtitle: "Data Simulation with Monte Carlo Methods"
author: "Marko Bachl"
institute: "University of Hohenheim"
format:
  revealjs:
      embed-resources: false
      height: 900
      width: 1600
      theme: [theme.scss]
      highlight-style: arrow-dark
      slide-number: c
      code-line-numbers: false
      history: false
      smaller: false
      title-slide-attributes:
        data-background-image: https://upload.wikimedia.org/wikipedia/commons/6/64/Uni_Hohenheim-Logo.svg
        data-background-position: "top right"
        data-background-size: auto
css: colors.css
execute:
  echo: true
---

```{r setup}
#| include: false
pacman::p_load(knitr, tictoc, tidyverse)
theme_set(hrbrthemes::theme_ipsum_rc(base_size = 25,
                                     axis_title_size = 25,
                                     strip_text_size = 20))
```

# Errors and power --- Torturing the *t*-test

## Traktandenliste

1)  Introduction & overview

2)  Monte Carlo Simulation?

3)  Proof by simulation --- The Central Limit Theorem (CLT)

4)  Errors and power --- Torturing the *t*-test

5)  Misclassification and bias --- Messages mismeasured

6)  Outlook: What's next?


## 4. Errors and power --- Torturing the *t*-test

a)  Comparison of Student's and Welch's *t*-tests

b)  *A priori* power calculation for Welch's *t*-test


## Student's *t*-test & Welch's *t*-test

Student's *t*-test for two independent means

$t = \frac{\bar{X}_1 - \bar{X}_2}{s_p \sqrt\frac{2}{n}}$, where $s_p = \sqrt{\frac{s_{X_1}^2+s_{X_2}^2}{2}}$ and $df = n_1 + n_2 − 2$

Welch's *t*-test for two independent means

$t = \frac{\bar{X}_1 - \bar{X}_2}{s_{\bar\Delta}}$, where ${\bar\Delta} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$ and $df = \frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)^2}{\frac{\left(s_1^2/n_1\right)^2}{n_1-1} + \frac{\left(s_2^2/n_2\right)^2}{n_2-1}}$

## Student's *t*-test & Welch's *t*-test

-   Old school advice: Student's *t*-test for equal variances, Welch's *t*-test for unequal variances.

    -   Higher power of Student's *t*-test if assumptions hold.

-   Modern advice: Always use Welch's *t*-test.

    -   Better if assumptions are violated; not worse if assumptions hold.

    -   e.g., Delacre, M., Lakens, D., & Leys, C. (2017). Why Psychologists Should by Default Use Welch's *t*-test Instead of Student's *t*-test. *International Review of Social Psychology, 30*(1). <https://doi.org/10.5334/irsp.82>

-   For those who don't care about *t*-tests: Idea also applies to heteroskedasticity-consistent standard errors.

## First Simulation study: False discoveries {.smaller}

1.  Question: What is the goal of the simulation?

    -   Comparison of Student's and Welch's *t*-tests

2.  Quantities of interest: What is measured in the simulation?

    -   $\alpha$ error rate of the tests

3.  Evaluation strategy: How are the quantities assessed?

    -   Comparison of empirical distribution of *p*-values with theoretical distribution under the Null hypothesis

4.  Conditions: Which characteristics of the data generating model will be varied?

    -   Group size ratio; Standard deviation ratio

5.  Data generating model: How are the data simulated?

    -   Random draws from normal distributions with equal means, but different group sizes and standard deviations

## Practical considerations

-   How does the `t.test()` work in *R*?

-   Which parameters and functions do we need for generating the data?

-   How do we implement the experimental design for the simulation study?

-   How do we run the simulation?

-   How do we assess and visualize the results?

## One simulation: Parameters

-   Two factors:

    -   Group sizes: Equal or unequal?

    -   Group standard deviations: Equal or unequal?

-   Implementation with ratios and fixed total sample size.

```{r One simulation: Parameters}
set.seed(28)
n = 150 # fixed total sample size
GR = 2 # n2/n1
n1 = round(n / (GR + 1))
n2 = round(n1 * GR)
sd1 = 1
SDR = 2  # sd2/sd1
sd2 = sd1 * SDR
```

## One simulation: *t*-tests

```{r One simulation: *t*-tests}
#| layout-ncol: 2
g1 = rnorm(n = n1, mean = 0, sd = sd1)
g2 = rnorm(n = n2, mean = 0, sd = sd2)
t.test(g1, g2) # Welch (default)
t.test(g1, g2, var.equal = TRUE) # Student
```

## Inspect the output of `t.test()`

```{r Inspect the output of t.test()}
t.test(g1, g2) %>% str()
```

```{r Extract p-value from t.test()}
t.test(g1, g2)$p.value
```

## Wrap it into a function

```{r Wrap it into a function}
sim_ttest = function(GR = 1, SDR = 1, n = 150) {
  n1 = round(n / (GR + 1))
  n2 = round(n1 * GR)
  sd1 = 1
  sd2 = sd1 * SDR
  g1 = rnorm(n = n1, mean = 0, sd = sd1)
  g2 = rnorm(n = n2, mean = 0, sd = sd2)
  Welch = t.test(g1, g2)$p.value
  Student = t.test(g1, g2, var.equal = TRUE)$p.value
  res = lst(Welch, Student)
  return(res)
}
```

## Call the function once

```{r Call the function once}
sim_ttest(GR = 2, SDR = 0.5) # Implies n1 = 50, n2 = 100, sd1 = 1, sd2 = 0.5
```

## Setup of experimental conditions

-   [`expand_grid()`](https://tidyr.tidyverse.org/reference/expand_grid.html) is the tidy implementation of [`base::expand.grid()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/expand.grid.html).
-   Creates a data frame from all combinations of the supplied vectors.

```{r Setup of experimental conditions: expand_grid example}
expand_grid(a = 1:2, b = 3:4)
```

## Setup of experimental conditions

-   Two between-simulation factors: `GR` and `SDR`

```{r Setup of experimental conditions}
conditions = expand_grid(GR = c(1, 2),
                         SDR = c(0.5, 1, 2)) %>% 
  rowid_to_column(var = "condition")
conditions
```

## Setup of experimental conditions

```{r Setup of experimental conditions 2}
#| echo: false
conditions %>% 
  mutate(implies = str_glue("n1 = {round(150 / (GR + 1))}, n2 = {round(150 - 150 / (GR + 1))}, sd1 = 1, sd2 = {1 * SDR}")) %>% 
  knitr::kable()
```

## Run simulation experiment

```{r Run simulation experiment}
#| cache: true
set.seed(689)
i = 10000 # number of sim runs per condition
tic() # simple way to measure wall time
sims = map_dfr(1:i, ~ conditions) %>% # each condition i times
  rowid_to_column(var = "sim") %>% # within simulation comparison
  rowwise() %>%
  mutate(p.value = list(sim_ttest(GR = GR, SDR = SDR))) %>% 
  unnest_longer(p.value, indices_to = "method")
toc() # simple way to measure wall time
```

## Inspect simulated data

```{r Inspect simulated data}
#| echo: false
sims
```

## Results: Histogram of *p*-values

  -   Distribution of *p*-values under the Null hypothesis should be $\sf{Uniform}(0, 1)$.

```{r Results: Histogram of p-values}
#| output-location: slide
#| fig-width: 16
#| fig-height: 8
sims %>% 
  ggplot(aes(p.value)) +
  geom_histogram(binwidth = 0.05, boundary = 0) + 
  facet_grid(method ~ GR + SDR, labeller = label_both) + 
  scale_x_continuous(breaks = c(1, 3)/4)
```

## Results: Q-Q-plot of *p*-values

-   Distribution of *p*-values under the Null hypothesis should be $\sf{Uniform}(0, 1)$.

```{r Results: Q-Q-plot of p-values}
#| output-location: slide
#| fig-width: 16
#| fig-height: 8
sims %>% 
  ggplot(aes(sample = p.value, color = method)) +
  geom_qq(distribution = stats::qunif, size = 0.2) + 
  geom_abline(slope = 1) + 
  facet_grid(GR ~ SDR, labeller = label_both)
```

## Results: Looking at the relevant tail {.smaller}
```{r Results: Looking at the relevant tail}
#| output-location: column
sims %>% 
  group_by(GR, SDR, method) %>% 
  summarise(P_p001 = mean(p.value <= 0.001),
            P_p01 = mean(p.value <= 0.01), 
            P_p05 = mean(p.value <= 0.05)) %>% 
  kable(digits = 3)
```


# Questions?

# Group exercise 2

## Exercise 2: Welch's *t*-test for count data

### Part 1: False discovery rates

- It is generally recommended to model count outcomes (e.g., number of likes of social media posts) with *Poisson* models (or *negative binomial* models, but that's for another simulation).

- However, if we only want to compare two groups on a count outcome, Welch's *t*-test might be an easier option and should also do well:
    - The mean is a consistent estimator of the rate parameter (CLT; see exercise 1a).
    - Welch's *t*-test is robust if the group standard deviations are unequal, which, by definition, would be the case if the counted event occurred at different rates in the two groups.

## Exercise 2a: False discovery rates

- We can test this assumption by simulation, similar to the comparison of Student's and Welch's *t*-test. More specifically, we first want to compare the false discovery rates of 
    - Student's *t*-test,
    - Welch's *t*-test, and
    - Poisson regression with a dummy predictor

- if the true data generating process is drawing from a Poisson distribution.

- Extend and adapt the previous simulation.

## Exercise 2a: Help {.scrollable}

### Expand if you need help.

### Don't expand if you want to figure it out by yourself. 

### Scroll down if you need more help on later steps.

Sample from a Poisson distribution
```{r Sample from a Poisson distribution}
#| eval: false
#| code-fold: true
rpois(n = n1, lambda = lambda1)
```

Poisson regression
```{r Poisson regression}
#| eval: false
#| code-fold: true
glm(outcome ~ group, data = data, family = poisson)
```

Data format for Poisson regression
```{r Data format for Poisson regression}
#| eval: false
#| code-fold: true
data.frame(outcome = c(g1, g2),
           group = c(rep("g1", n1), rep("g2", n2)))
```

Extract *p*-value from fitted Poisson regression object
```{r Extract p-value from fitted Poisson regression object}
#| eval: false
#| code-fold: true
GLM = glm(outcome ~ group, data = data, family = poisson)
coef(summary(GLM))[2, 4]
```

Complete simulation function
```{r Complete simulation function}
#| eval: false
#| code-fold: true
# No standard deviation ratio any more because SD fixed at sqrt(lambda)
# M_diff not necessary at this stage, but useful later
sim_ttest_glm = function(n = 200, GR = 1, lambda1 = 1, M_diff = 0) {
  n1 = round(n / (GR + 1))
  n2 = round(n1 * GR)
  lambda2 = lambda1 + M_diff
  g1 = rpois(n = n1, lambda = lambda1)
  g2 = rpois(n = n2, lambda = lambda2)
  Welch = t.test(g1, g2)$p.value
  Student = t.test(g1, g2, var.equal = TRUE)$p.value
  GLM = glm(outcome ~ group,
            data = data.frame(outcome = c(g1, g2),
                              group = c(rep("g1", n1), rep("g2", n2))),
            family = poisson)
  res = lst(Welch, Student, GLM = coef(summary(GLM))[2, 4])
  return(res)
}
```

Conditions
```{r Conditions}
#| eval: false
#| code-fold: true
# It makes sense to vary group size ratio and lambda
conditions = expand_grid(GR = c(0.5, 1, 2), 
                         n = 600,
                         lambda1 = c(1, 10, 20),
                         M_diff = 0) %>% 
  rowid_to_column(var = "condition")
```

Run simulation
```{r Run simulation}
#| eval: false
#| code-fold: true
# use smaller i for now because additional glm() needs time
set.seed(35)
i = 1000
tic()
sims = map_dfr(1:i, ~ conditions) %>%
  rowid_to_column(var = "sim") %>%
  rowwise() %>%
  mutate(p.value = list(sim_ttest_glm(n = n, M_diff = M_diff,
                                 GR = GR, lambda1 = lambda1))) %>% 
  unnest_longer(p.value, indices_to = "method")
toc()
```

Plots
```{r Plots}
#| eval: false
#| code-fold: true
# Histogram of p-values
sims %>% 
  ggplot(aes(p.value)) +
  geom_histogram(binwidth = 0.1, boundary = 0) + 
  facet_grid(method ~ lambda1 + GR, labeller = label_both) + 
  scale_x_continuous(breaks = c(1, 3)/4)
# QQ-Plot of p-values
sims %>% 
  ggplot(aes(sample = p.value, color = method)) +
  geom_qq(distribution = stats::qunif, geom = "line") + 
  geom_abline(slope = 1) + 
  facet_grid(GR ~ lambda1, labeller = label_both)
# Rates of p < .05
sims %>% 
  group_by(GR, lambda1, method) %>% 
  summarise(P_p05 = mean(p.value < 0.05)) %>% 
  ggplot(aes(factor(GR), P_p05, color = method, group = method)) + 
  geom_point() + geom_line() + 
  geom_hline(yintercept = 0.05, linetype = 2) + 
  facet_wrap(vars(lambda1), labeller = label_both) +
  labs(x = "Group size ratio", y = "Proportion p < 0.05")
```
